---
title: "In-class Exercise 1: First Date with Geospatial Data Analytics"
date: "18 November 2023"
date-modified: "last-modified"
---

## Getting Started

the code chunk below load the following packages:

-   tmap: thematic mapping.

-   sf: for geospatial data handling.

-   tidyverse: for non-spatial data handling.

```{r}
pacman::p_load(tmap, sf, tidyverse, knitr)
```

## Importing Data

### Importing OD data

import data of *Passenger Volume by Origin Destination Bus Stop* dataset from LTA Data Mall utilising read_csv() of **readr** package.

```{r}
odbus = read_csv("data/aspatial/origin_destination_bus_202308.csv")
```

```{r}
glimpse(odbus)
```

changing origin and destination pt code from CHR to Factor

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```

```{r}
glimpse(odbus)
```

picking data from 7 to 9 and in weekday

```{r}
origin7_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 7 &
           TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

```{r}
kable(head(origin7_9))
```

```{r}
write_rds(origin7_9, "data/rds/origin7_9.rds")
```

```{r}
origin7_9 <- read_rds("data/rds/origin7_9.rds")
```

### Importing geospatial data

#### bus stop

```{r}
busstop <- st_read(dsn = "data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(busstop)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                   layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
mpsz
```

```{r}
glimpse(mpsz)
```

## **Geospatial data wrangling**

### **Combining Busstop and mpsz**

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
```

```{r}
write_rds(busstop_mpsz, "data/rds/busstop_mpsz.csv")  
```

```{r}
origin_data <- left_join(origin7_9 , busstop_mpsz,
            by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = SUBZONE_C)
```

```{r}
duplicate <- origin_data %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
```

```{r}
origin_data <- unique(origin_data)
```

```{r}
mpsz_origtrip <- left_join(mpsz, 
                           origin_data,
                           by = c("SUBZONE_C" = "ORIGIN_SZ"))
```

```{r}
tm_shape(mpsz_origtrip)+
  tm_fill("TRIPS", 
          style = "quantile", 
          palette = "Blues",
          title = "Passenger trips") +
  tm_layout(main.title = "Passenger trips generated at planning sub-zone level",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
